---
title: "Virginia Schools Pre- & Post-Pandemic"
author: "Cassandra Sperow"
date: '2022-06-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploration of Pre- & Post-Pandemic Pass Rate Scores from the Standards of Learning (SOL) Test Results of Schools in the State of Virginia, USA

- A previous version of sorts for Fairfax County is also available in this repo. 

- This analysis is for my own curiosity and should not be construed as anything extensive in regards to judging schools or judging how good a school may be. There are too many factors and circumstances that go into what makes a good school. 


## Original public data source:

[Virginia Department of Education SOL Test Results](https://www.doe.virginia.gov/statistics_reports/sol-pass-rates/index.shtml)





### Import Libraries
```{r}
library(tidyverse)
```


### Import Data
```{r}
schools <- read_csv("../data/2020-2021-school-test-by-test.csv", skip =2)

schools %>% head
```

### Checking columns where pass rate should have been a number

```{r}
# There is a "<" character that needs to be counted as NA
schools$`2018-2019 Pass Rate` %>% unique()
```


### Mutating pass rate columns into number columns with new NA definition to include the character "<"

- Also changing the school type column to a categorical variable and renaming it in one mutate()
```{r}
# setting the NA values to include "<" because there was not a value given or key for what the "<" indicates 
schools %>% 
  mutate(sch_type = parse_factor(`Sch Type`)) %>% 
  mutate(pass_rate_18_19 = parse_integer(`2018-2019 Pass Rate`, 
                                         na = c("", "NA", NA, "<"), 
                                         trim_ws = TRUE), 
         pass_rate_20_21 = parse_integer(`2020-2021 Pass Rate`, 
                                         na = c("", "NA", NA, "<"), 
                                         trim_ws = TRUE),
         county = `Div Name`) -> schools
  
# checking to make sure they're all numbers
filter(schools, is.character(pass_rate_20_21))
```
### Calculating a Percent Change between school year 2018-2019 and 2020-2021. 

Note: Schools in Virginia did not post any data for the school year 2019-2020. 

```{r}
schools %>% 
  # selecting necessary columns rather than whole dataset
  select(county, sch_type, `Sch Name`, Test, Grade, pass_rate_18_19, pass_rate_20_21) %>% 
  # calculating point difference between pre- and post-pandemic
  mutate(point_diff =  pass_rate_20_21 - pass_rate_18_19) %>% 
  # creating percent change column
  mutate(pct_change = (point_diff/pass_rate_18_19)*100) %>% 
  # changing into categorical variable
  mutate(Test = parse_factor(Test)) -> schools

schools %>% glimpse()
```

## Basic Descriptive Stats

- How many unique schools?
```{r}
schools %>% 
  summarise(`Sch Name`) %>%  
  unique() %>% 
  nrow
```

- How many county/city school districts within the State of Virginia?
```{r}
schools %>% 
  summarise(county) %>% 
  unique() %>% 
  nrow()
```

- How many schools per district? Which one has the most schools (ie, the largest)?
```{r}
schools %>% 
  group_by(county, `Sch Name`) %>% 
  summarise(county, `Sch Name`) %>% 
  unique() %>% 
  group_by(county) %>% 
  summarise(count = n()) %>% 
  arrange(-count)
```



- What are the categories of schools?
```{r}
schools %>% 
  summarise(sch_type) %>%  
  unique()
```


- What are the Test categories?

```{r}
schools$Test %>% levels()
```


### Distribution of 2018-2019 Pass Rates for All Schools
```{r}
schools %>% 
  ggplot(aes(x = pass_rate_18_19, y = Test)) +
  geom_boxplot(outlier.shape = "triangle", outlier.color = "red") + 
  theme_bw() +
  stat_summary(fun = mean, color = "red") + 
  ggtitle("Pass Rates per Subject Area for Schools in Virginia, USA", 
          subtitle = "for the school year 2018-2019")
```

Key: 

- The triangles are considered outliers. 
- The circles represent the mean (average) pass rate for that test subject area for the whole State of Virginia.
- The black lines inside each boxplot indicate the median pass rate.


